---
title: "p8105_hw3_tq2171"
author: "Tingyu Qian"
date: "2024-10-9"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Problem 1
```{r}
# load the library and the dataset
library(p8105.datasets)
data("ny_noaa")
```

In the dataset "ny_noaa", it contains `r nrow(ny_noaa)` and `r ncol(ny_noaa)` columns, which means there are 7 variables. Name of the 7 variables are the following: `r names(ny_noaa)`.

The following are the variables introduction:

id: Weather station ID

date: Date of observation

prcp: Precipitation (tenths of mm)

snow: Snowfall (mm)

snwd: Snow depth (mm)

tmax: Maximum temperature (tenths of degrees C)

tmin: Minimum temperature (tenths of degrees C)

The type of each variable are the following: id is `r class(ny_noaa$id)`,  date is `r class(ny_noaa$date)`, prcp is `r class(ny_noaa$prcp)`, snow is `r class(ny_noaa$snow)`, snwd is `r class(ny_noaa$snwd)`, tmax is `r class(ny_noaa$tmax)`, tmin is `r class(ny_noaa$tmin)`. Variables "prcp", "snow", "snwd" are the integer variables. Variables "id", "tmax" and "tmin" are character variables. There are `r sum(is.na(ny_noaa$prcp))` or `r round((sum(is.na(ny_noaa$prcp))/2595176)*100, 2)`% missing value in the variable "prcp", `r sum(is.na(ny_noaa$snow))` or `r round((sum(is.na(ny_noaa$snow))/2595176)*100, 2)`% missing value in the variable "snow", `r sum(is.na(ny_noaa$snwd))` or `r round((sum(is.na(ny_noaa$snwd))/2595176)*100, 2)`% missing value in the variable "snwd", `r sum(is.na(ny_noaa$tmax))` or `r round((sum(is.na(ny_noaa$tmax))/2595176)*100, 2)`% missing value in the variable "tmax" and `r sum(is.na(ny_noaa$tmin))`  or `r round((sum(is.na(ny_noaa$tmin))/2595176)*100, 2)`% missing value in the variable "tmin". 